Solving the grand challenge using an algorithmic an distributed approach

### 2015-03-16 paper/query_implementation.tex
adding order list data structure

\begin{table}
\centering
\caption{Operations of \textit{NodeList} Data Structure}
\begin{tabular}{|l|l|} \hline
Operation Signature & Time Complexity \\ \hline \hline
add(key : Object, value : NodeValue) & O(log(n)) \\ \hline 
get(key : Object) : NodeValue & O(1) \\ \hline
remove(key : Object) : NodeValue & O(log(n)) \\ \hline
decrementPosition(key : Object) & O(log(n)) \\ \hline
incrementPosition(key : Object) & O(log(n)) \\ \hline
getTopValues() : List<NodeValue> & O(1) \\ \hline
\end{tabular}
\end{table}

>>>>

\begin{table}
\centering
\caption{Operations of \textit{NodeList} Data Structure}
\begin{tabular}{|l|l|} \hline
Operation Signature & Time Complexity \\ \hline \hline
add(key : Object, value : NodeValue) & O(log(n)) \\ \hline 
get(key : Object) : NodeValue & O(1) \\ \hline
remove(key : Object) : NodeValue & O(log(n)) \\ \hline
decrementPosition(key : Object) & O(log(n)) \\ \hline
incrementPosition(key : Object) & O(log(n)) \\ \hline
getTopValues() : List<NodeValue> & O(1) \\ \hline
\end{tabular}
\label{nodelist_api}
\end{table}

Table \ref{nodelist_api} shows the operations of this data structure. The NodeValue is an interface to plug any user defined type as a value. For an example, the route query can use a RouteCount object to keep route count while profitable cells query can use CellProfict object to keep fare details. This interface contains a method called compare to define the order of the values. The NodeList data structure provides the notion of a sorted values starting from position 0 (highest value) to its users. The decrementPostion operation moves the corresponding value towards position 0 until the correct position according to sorted order. Therefore this method should be called after increasing a value of an existing object. Similarly incrementPosition operation can be used to move a value to correct position after decreasing the value of an existing object. GetTopValues operation returns the top ten values of the list. Query algorithm sections further elaborate usage of these methods.

One way of implementing this data structure is to use a heap. Heap supports adding and extracting maximum value operations in O(log(n)) time. However since this application requires top 10 values, we need to retrieve maximum value ten times and insert them back, if we just a heap. On the other hand if we use a doubly linked list to keep values, it will take O(n) time for  decrementPostion and   incrementPosition operations. Therefore we can use a doubly linked list to keep first 10 values (this makes time complexity for all operations with this data structure O(1)) and heap to keep other values to avoid above problems.  A Map which points to values can be used to retrieve values using a key.

NodeList data structure uses two data structures called OrderedList to keep first 10 values and DynamicHeap to keep other values. In this way it has to change only the last value of OrderedList and maximum value of Dynamic heap as required. Following sections describes how these structures implemented with their operations.

\subsection{OrderedList data structure}

\begin{table}
\centering
\caption{Operations of \textit{OrderedList} Data Structure}
\begin{tabular}{|l|l|} \hline
Operation Signature & Time Complexity \\ \hline \hline
add(key : Object, value : NodeValue) & O(n) \\ \hline
containsKey(key : Object) : boolean & O(1) \\ \hline
get(key : Object) : NodeValue & O(1) \\ \hline
incrementPosition(key : Object) & O(n) \\ \hline
decrementPosition(key : Object) & O(n) \\ \hline
remove(key : Object) : NodeValue & O(1) \\ \hline
getLastPosition() : int & O(1) \\ \hline
getLastKey() : Object & O(1) \\ \hline
\end{tabular}
\label{orderedlist_api}
\end{table}

Table \ref{orderedlist_api} shows the operations of the \textit{OrderedList} data structure. First six methods has the same meanings as the \textit{NodeList} data structure and last two can be used to get the size of the list and the value of the last object. We internally keep a counter and a pointer to tail to support these operations in O(1) time. Figure \ref{orderedlist_impl} shows the implementation details. The Map structure (an HashMap) is used to retrieve object pointers with O(1) expected time. Each node of the doubly linked list points to previous and next nodes while storing the \textit{NodeValue}.

\begin{figure}[!t]
        \centering
        \includegraphics[width=3.0in]{orderedlist.png}
        \caption{\textit{OrderedList} data structure implementation}
        \label{orderedlist_impl}
\end{figure}
